Evaluating FAIR Digital Object as a distributed object system

### 2023-02-02 latex/table5.tex
all the citations. Bioschemas by article.

\caption{Assessing RDA's FAIR Data Maturity Model \cite{1GrKTFaK2,UzQhqk0M} (first 2 columns) against the FDO guidelines \cite{RwvirqWg}, FDO implemented with the protocol DOIPv2 \cite{13TcbsZF6}, Linked Data Platform (LDP) \cite{7szz7dwO} and examples from Linked Data practices in general. (--- indicates \emph{Unspecified}, may be possible with additional conventions)

@@

RDA-F4-01M & Metadata is offered in such a way that it can be harvested and indexed & FDOF10 & No, registries not required (except Data Type Registries). Handle registry only searchable by PID. & Not specified & Not specified, several registries/catalogues for vocabularies/types (e.g. \cite{1Gsq23e51}). Indexing by search engines if exposing HTML w/schema.org. \\

@@

RDA-A1-04D & Data is accessible through standardised protocol & G9 & (see above) & HTTP \cite{8HJqcF1Q} & HTTP/HTTPS, FTP (now less common), GridFTP \cite{18LsXvKGJ} (for large data), ARK \cite{voDQ6diM} \\

@@

RDA-A1.1-01M & Metadata is accessible through a free access protocol & G1 G8 G9 & Partially realised: Handle system is open\footnote{The \texttt{Handle.net} system was previously covered by software patent \href{https://patents.google.com/patent/US6135646A/en}{US6135646A} which \href{https://circleid.com/posts/20161025_selling_dona_snake_oil_at_the_itu\#11461}{expired} in 2013.} protocol \cite{I9Kx0Hjy}. One server implementation \cite{1CvYRVhy2}, free\footnote{The \href{http://www.handle.net/HNRj/HNR-9-License.pdf}{Handle.net public license}{]} is not OSI-approved \cite{xVJMmGVr}  as an open source license -- it includes usage restrictions and requires Service Agreements. It is not a DOIP requirement to host a local Handle instance, e.g.~EOSC provides the \href{https://sp.eudat.eu/catalog/resources/fc6b2d30-09cd-4c25-b71a-7bc6de77910c}{B2HANDLE} service for acquiring Handle prefixes.}. One DOIPv2 implementation (\href{https://www.cordra.org/}{CORDRA}): free under BSD-like license (not recognised as Open Source). & \href{http://www.w3.org/TR/ldp/}{LDP} is open W3C recommendation. \href{https://www.w3.org/wiki/LDP_Implementations}{Multiple LDP implementations}. & DNS, HTTP, TLS, RDF standards are open, free and universal, large number of Open Source clients and \href{https://en.wikipedia.org/wiki/Comparison_of_web_server_software}{servers}. \\

@@

RDA-I2-01D & Data uses FAIR-compliant vocabularies & --- & --- & --- & Uncommon, except for some XML and RDF-embedding formats, e.g.~Extensible Metadata Platform (XMP) \cite{1BlxdxOwa} \\
RDA-I3-01M & Metadata includes references to other metadata & FDOR8 & Implied (attributes to PIDs), currently unspecified if given attribute is value or reference & --- & By definition (Linked Data reference existing URIs \cite{o4h92LpS}), \texttt{rdfs:seeAlso}, FAIR signposting \cite{snykkm7R} \texttt{describedby} \\

@@

RDA-I3-03M & Metadata includes qualified references to other metadata & (FDOR3) & Qualification by attribute keys defined per FDO Type & \href{https://www.w3.org/TR/ldp/\#dfn-linked-data-platform-direct-container}{LDP Direct Container} & Qualifications by property, PROV bundles \cite{QpDmajc4}, \href{https://schema.org/Role}{schema.org/Role} \\
RDA-I3-04M & Metadata include qualified references to other data & (FDOR3) & Qualification by attribute keys defined per FDO type & \href{https://www.w3.org/TR/ldp/\#dfn-linked-data-platform-indirect-container}{LDP Indirect Container} & Qualifications by property, n-ary indirection (schema.org Role \cite{10HA0mMBx}, \texttt{prov:specializationOf} \cite{ZUgwvAHE}, OAI-ORE Proxy \cite{ID1pTGry}) \\
RDA-R1-01M & Plurality of accurate and relevant attributes are provided to allow reuse & FDOF4 & Required. Kernel metadata attributes desired \cite{Fi62cJAo} but not assigned PIDs yet. & Unspecified. Multiple metadata records can allow multiple semantic profiles. & Large number of general and domain-specific vocabularies can make it hard to find relevant attributes. Rough consensus on kernel metadata: schema.org \cite{10pzKCEUE}, Dublin Core Terms \cite{wAbZpOsr}, DCAT \cite{lnH0hEuh}, FOAF \cite{XmUY0cZB} \\
RDA-R1.1-01M & Metadata includes information about the licence under which the data can be reused & --- & \texttt{licenseConditions} URL/PID in kernel metadata \cite{Fi62cJAo} & --- & Dublin Core Terms \texttt{dct:license} frequently recommended, frequently not required, e.g.~\href{https://www.w3.org/TR/vocab-dcat-2/\#Property:distribution_license}{by DCAT 2} \cite{lnH0hEuh} \\

@@

RDA-R1.2-02M & Metadata includes provenance information according to a cross-community language & FDOR9 FDOR8 & --- & --- & W3C PROV-O \cite{9T4j3N4e}, PAV \cite{1E4OfHyhI}, Dublin Core Terms \cite{hKblceo6} \\
RDA-R1.3-01M & Metadata complies with a community standard & FDOR10 FROR8 & (Emerging, e.g.~DiSSCo Digital Specimen \cite{zZTtJruI}) & --- & Common, e.g.~DCAT 2 \cite{zt8AAjcZ}, BioSchemas \cite{1F8S9FeHf} \\

@@

RDA-R1.3-02M & Metadata is expressed in compliance with a machine-understandable community standard & FDOF4 FDOF10 & Recommended & --- & Common practice for ontologies, specially in bioinformatics, e.g.~BioPortal \cite{1Gsq23e51}, Darwin Core \cite{9jaVzCbL} \\

>>>>

\caption{Assessing RDA's FAIR Data Maturity Model \autocite{1GrKTFaK2,UzQhqk0M} (first 2 columns) against the FDO guidelines \autocite{RwvirqWg}, FDO implemented with the protocol DOIPv2 \autocite{13TcbsZF6}, Linked Data Platform (LDP) \autocite{7szz7dwO} and examples from Linked Data practices in general. (--- indicates \emph{Unspecified}, may be possible with additional conventions)

@@

RDA-F4-01M & Metadata is offered in such a way that it can be harvested and indexed & FDOF10 & No, registries not required (except Data Type Registries). Handle registry only searchable by PID. & Not specified & Not specified, several registries/catalogues for vocabularies/types (e.g. \autocite{NCBOBioPortal}). Indexing by search engines if exposing HTML w/schema.org. \\

@@

RDA-A1-04D & Data is accessible through standardised protocol & G9 & (see above) & HTTP \autocite{rfc9110} & HTTP/HTTPS, FTP (now less common), GridFTP \autocite{allcockGlobusStripedGridFTP} (for large data), ARK \autocite{ARKIdentifierScheme} \\

@@

RDA-A1.1-01M & Metadata is accessible through a free access protocol & G1 G8 G9 & Partially realised: Handle system is open\footnote{The \texttt{Handle.net} system was previously covered by software patent \href{https://patents.google.com/patent/US6135646A/en}{US6135646A} which \href{https://circleid.com/posts/20161025_selling_dona_snake_oil_at_the_itu\#11461}{expired} in 2013.} protocol \autocite{rfc3652}. One server implementation \autocite{HandleNetRegistry}, free\footnote{The \href{http://www.handle.net/HNRj/HNR-9-License.pdf}{Handle.net public license} is not OSI-approved \autocite{LicensesStandardsOpen}  as an open source license -- it includes usage restrictions and requires Service Agreements. It is not a DOIP requirement to host a local Handle instance, e.g.~EOSC provides the \href{https://sp.eudat.eu/catalog/resources/fc6b2d30-09cd-4c25-b71a-7bc6de77910c}{B2HANDLE} service for acquiring Handle prefixes.}. One DOIPv2 implementation (\href{https://www.cordra.org/}{CORDRA}): free under BSD-like license (not recognised as Open Source). & \LDP is open W3C recommendation \autocite{w3-ldp}. \href{https://www.w3.org/wiki/LDP_Implementations}{Multiple LDP implementations}. & DNS, HTTP, TLS, RDF standards are open, free and universal, large number of Open Source clients and \href{https://en.wikipedia.org/wiki/Comparison_of_web_server_software}{servers}. \\

@@

RDA-I2-01D & Data uses FAIR-compliant vocabularies & --- & --- & --- & Uncommon, except for some XML and RDF-embedding formats, e.g.~Extensible Metadata Platform (XMP) \autocite{iso16684} \\
RDA-I3-01M & Metadata includes references to other metadata & FDOR8 & Implied (attributes to PIDs), currently unspecified if given attribute is value or reference & --- & By definition (Linked Data reference existing URIs \autocite{DataW3C}), \texttt{rdfs:seeAlso}, FAIR signposting \autocite{vandesompelFAIRSignpostingProfile2022} \texttt{describedby} \\

@@

RDA-I3-03M & Metadata includes qualified references to other metadata & (FDOR3) & Qualification by attribute keys defined per FDO Type & \href{https://www.w3.org/TR/ldp/\#dfn-linked-data-platform-direct-container}{LDP Direct Container} & Qualifications by property, PROV bundles \autocite{w3-prov-links}, \href{https://schema.org/Role}{schema.org/Role} \\
RDA-I3-04M & Metadata include qualified references to other data & (FDOR3) & Qualification by attribute keys defined per FDO type & \href{https://www.w3.org/TR/ldp/\#dfn-linked-data-platform-indirect-container}{LDP Indirect Container} & Qualifications by property, n-ary indirection (schema.org Role \autocite{hollandIntroducingRole2014}, \texttt{prov:specializationOf} \autocite{w3-prov-o}, OAI-ORE Proxy \autocite{ORESpecificationAbstract}) \\
RDA-R1-01M & Plurality of accurate and relevant attributes are provided to allow reuse & FDOF4 & Required. Kernel metadata attributes desired \autocite{fdo-KernelAttributes} but not assigned PIDs yet. & Unspecified. Multiple metadata records can allow multiple semantic profiles. & Large number of general and domain-specific vocabularies can make it hard to find relevant attributes. Rough consensus on kernel metadata: schema.org \autocite{10pzKCEUE}, Dublin Core Terms \autocite{DCMIMetadataTerms}, DCAT \autocite{w3-vocab-dcat-2}, FOAF \autocite{FOAFVocabularySpecification} \\
RDA-R1.1-01M & Metadata includes information about the licence under which the data can be reused & --- & \texttt{licenseConditions} URL/PID in kernel metadata \autocite{fdo-KernelAttributes} & --- & Dublin Core Terms \texttt{dct:license} frequently recommended, frequently not required, e.g.~\href{https://www.w3.org/TR/vocab-dcat-2/\#Property:distribution_license}{by DCAT 2} \autocite{w3-vocab-dcat-2} \\

@@

RDA-R1.2-02M & Metadata includes provenance information according to a cross-community language & FDOR9 FDOR8 & --- & --- & W3C PROV-O \autocite{w3-prov-o}, PAV \autocite{ciccaresePAVOntologyProvenance2013e}, Dublin Core Terms \autocite{DCMIMetadataTerms} \\
RDA-R1.3-01M & Metadata complies with a community standard & FDOR10 FROR8 & (Emerging, e.g.~DiSSCo Digital Specimen \autocite{hardistySpecimenDataRefinery2022a}) & --- & Common, e.g.~DCAT 2 \autocite{w3-vocab-dcat-2}, BioSchemas \autocite{BioschemasBioschemas} \\

@@

RDA-R1.3-02M & Metadata is expressed in compliance with a machine-understandable community standard & FDOF4 FDOF10 & Recommended & --- & Common practice for ontologies, specially in bioinformatics, e.g.~BioPortal \autocite{NCBOBioPortal}, Darwin Core \autocite{wieczorekDarwinCoreEvolving2012} \\
